Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-25 appendix.tex
Many small changes to all files -- GB final edits

which is the value of $\Delta Q_0$ that the other slopes ($\Delta
Q_1$, $\Delta Q_2$, and $\Delta Q_3$) imply, normalized by the desired
value of $\Delta Q_0$ itself.  If $0 < s < 1$, then no adjustment
needs to be made, as the monotonicity constraints are met.  If these
equalities are not met, then, if $s<0$ we define $\chi_n = 1$ if
$\Delta Q_n/\Delta Q_0 < 0$, and 0 otherwise (if $s > 1$, this is
reversed, so $\chi_n = 1$ if $\Delta Q_n/\Delta Q_0 > 0$, and 0
otherwise).  These weights are used to determine which slopes to
modify to match the $\Delta Q_0$ constraint.  We then compute the
amount of adjustment required:

@@

This interpolation method provides a third-order accuracy based on the
Triangular Shaped Cloud (TSC) methodology \citep{Hockney88}.  As
usual, in one dimension, we define the parent values as $Q_{-1}$,
$Q_0$, and $Q_{+1}$, where the central parent cell has a left edge at
$x_0$ and width $\Delta x$.  Then, the interpolated value for a
subgrid cell $q_i$ with a cell left edges at $x_i = x_0 + i \Delta
x^p/r$ where $i$ runs from 0 to $r-1$ for refinement factor $r$, is
given by:

@@

Finally, for completeness, we include a first-order accurate piecewise
constant interpolator, for which, using the same definitions in the
previous case, we take $q_i = Q_0$, for $i = 0$ to $r-1$.

>>>>

which is the value of $\Delta Q_0$ that the other slopes ($\Delta
Q_1$, $\Delta Q_2$, and $\Delta Q_3$) imply, normalized by the desired
value of $\Delta Q_0$ itself.  If $0 < s < 1$, then no adjustment
needs to be made, as the monotonicity constraints are met.  If these
equalities are not met, then, if $s<0$ we define $\chi_n = 1$ if
$\Delta Q_n/\Delta Q_0 < 0$, and $\chi_n = 0$ otherwise (if $s > 1$, this is
reversed, so $\chi_n = 1$ if $\Delta Q_n/\Delta Q_0 > 0$, and 0
otherwise).  These weights are used to determine which slopes to
modify to match the $\Delta Q_0$ constraint.  We then compute the
amount of adjustment required:

@@

This interpolation method provides third-order accuracy based on the
Triangular Shaped Cloud (TSC) methodology \citep{Hockney88}.  As
usual, in one dimension, we define the parent values as $Q_{-1}$,
$Q_0$, and $Q_{+1}$, where the central parent cell has a left edge at
$x_0$ and width $\Delta x$.  Then, the interpolated value for a
subgrid cell $q_i$ with a cell left edges at $x_i = x_0 + i \Delta
x^p/r$ where $i$ runs from 0 to $r-1$ for refinement factor $r$, is
given by:

@@

Finally, for completeness, we include a first-order accurate piecewise
constant interpolator for which, using the same definitions as in the
previous case, we take $q_i = Q_0$, for $i = 0$ to $r-1$.

### 2013-06-25 introduction.tex
Many small changes to all files -- GB final edits

Due to the high spatial and temporal dynamical ranges involved,
astrophysical and cosmological phenomena present a taxing challenge for
simulators. To tackle such situations, a number of numerical
techniques have been developed that can be broadly split into the
gridless, Lagrangian methods and grid-based, Eulerian schemes. The
most commonly used is an example of the first type known as Smoothed
Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achieved
much success, particularly in regimes dominated by gravity. However,
its development to include an increasing number of sought-after
physical processes is still at a relatively early stage when compared
with the effort put into the latter type of Eulerian grid-based
hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997,
Woodward84}.

@@

There have been a number of numerical methods described in the
astronomical literature that contain elements of SAMR or have a
similar aim. For example, the N-body solver developed by
\citet{Villumsen89} used non-adaptive meshing to increase the
resolution in pre-selected regions. This static approach was later
used extensively when applied to hydrodynamics
\citep[e.g.,][]{Ruffert94, Anninos94}. Adding adaptivity is a more
recent enhancement, and there are now a number of N-body codes that
possess this feature, both with and without hydrodynamics
\citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97,
Truelove98, flash_method, MLAPM01, Yahagi01, RAMSES, Quilis04,
Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}. Of these,
perhaps the most comparable and widely used are FLASH
\citep{flash_method}, which uses grid blocks of fixed size, and RAMSES
\citep{RAMSES} and ART \citep{ART97}, both of which refine individual
cells. It is also possible to deform the grid to obtain high
resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a
few codes have adopted an unstructured approach based on a moving
Voronoi mesh \citep{Arepo10, Tess11}.

In this paper, we present \enzo, a structured adaptive mesh refinement
(SAMR) code. Originally developed for cosmological hydrodynamics,
\enzo\ has since been used on a wide variety of problems.  It has
grown to become a general tool for astrophysical fluid dynamics and is
intended to be efficient, accurate and easily extended to include new
capabilities.  Although many of the components of the \enzo\ code have
been described in previous publications \citep{1995CoPhC..89..149B,
BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01,
Oshea04, 2007arXiv0705.1556N,
WangAbelZhang08,ReynoldsHayesPaschosNorman2009,Collins10,Wise11_Moray},
there has previously been no systematic and complete description of
the code.  In this paper we provide that description, filling in many
previous omissions and showing the code's performance for a wide
variety of test problems.

The \enzo\ code has been extensively used over the last two decades in
a wide variety of problems, resulting in the publication of more than
100 peer-reviewed papers based on results obtained with the code. The
variety of astrophysical systems that \enzo\ has been used for include
galaxies \citep{2003ApJ...587...13T, 2012MNRAS.425..641L,
  2013MNRAS.432.1989S}, galaxy
clusters \citep{Loken02, Xu11, Skillman13}, the interstellar medium
\citep{Slyz05}, the intergalactic medium \citep{Fang01, 
2011ApJ...731....6S}, the circumgalactic
medium~\citep{2013MNRAS.430.1548H}, cooling flows \citep{Li12, 2013ApJ...763...38S},
turbulence \citep{Kritsuk04, 2007ApJ...665..416K, 2009JPhCS.180a2020K, Collins11}, the
formation of the first stars \citep{ABN02, 2007ApJ...654...66O,
2009Sci...325..601T, Xu08}, and the formation of stars in our own Galaxy
\citep{Collins11, 2011ApJ...727L..20K, Collins12a}.

>>>>

Due to the high spatial and temporal dynamical ranges involved,
astrophysical and cosmological phenomena present a taxing challenge for
simulators. To tackle such situations, a number of numerical
techniques have been developed that can be broadly split into
gridless, Lagrangian methods and grid-based, Eulerian schemes. The
most commonly used is an example of the first type known as Smoothed
Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achieved
much success, particularly in regimes dominated by gravity. However,
its development to include an increasing number of sought-after
physical processes is still at a relatively early stage when compared
with the effort put into the latter type of Eulerian grid-based
hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997,
Woodward84}.

@@

There have been a number of numerical methods described in the
astronomical literature that contain elements of SAMR or have a
similar aim. For example, the N-body solver developed by
\citet{Villumsen89} used non-adaptive meshing to increase the
resolution in pre-selected regions. This static approach was later
used extensively when applied to hydrodynamics
\citep[e.g.,][]{Ruffert94, Anninos94}. Adding adaptivity is a more
recent enhancement, and there are now a number of codes that
possess this feature, both with and without hydrodynamics
\citep{Couchman91, Jessop94, Suisalu95, Splinter96, Gelato97, ART97,
Truelove98, flash_method, MLAPM01, Yahagi01, RAMSES, Quilis04,
Ziegler05, Zhang06, Astrobear09, Pluto-amr, GAMER, Nyx}. Of these,
perhaps the most comparable and widely used are FLASH
\citep{flash_method}, which uses grid blocks of fixed size, and RAMSES
\citep{RAMSES} and ART \citep{ART97}, both of which refine individual
cells. It is also possible to deform the grid to obtain high
resolution \citep[e.g.,][]{Gnedin95, Xu97, Pen98}, and more recently a
few codes have adopted an unstructured approach based on a moving
Voronoi mesh \citep{Arepo10, Tess11}.

In this paper, we present \enzo\ 2.3, a structured adaptive mesh refinement
(SAMR) code. Originally developed for cosmological hydrodynamics,
\enzo\ has since been used on a wide variety of problems.  It has
grown to become a general tool for astrophysical fluid dynamics and is
intended to be efficient, accurate and easily extended to include new
capabilities.  Although many of the components of the \enzo\ code have
been described in previous publications \citep{1995CoPhC..89..149B,
BryanThesis96, Bryan97a, Bryan97b, Norman99, BryanCompSci99, Bryan01,
Oshea04, 2007arXiv0705.1556N,
WangAbelZhang08,ReynoldsHayesPaschosNorman2009,Collins10,Wise11_Moray},
there has previously been no systematic and complete description of
the code.  In this paper we provide that description, filling in many
previous omissions and showing the code's performance for a wide
variety of test problems.

The \enzo\ code has been extensively used over the last two decades in
a wide variety of problems, resulting in the publication of more than
100 peer-reviewed papers. The
variety of astrophysical systems that \enzo\ has been used for include
galaxies \citep{2003ApJ...587...13T, 2012MNRAS.425..641L,
  2013MNRAS.432.1989S}, galaxy
clusters \citep{Loken02, Xu11, Skillman13}, the interstellar medium
\citep{Slyz05}, the intergalactic medium \citep{Fang01, 
2011ApJ...731....6S}, the circumgalactic
medium~\citep{2013MNRAS.430.1548H}, cooling flows \citep{Li12, 2013ApJ...763...38S},
turbulence \citep{Kritsuk04, 2007ApJ...665..416K, 2009JPhCS.180a2020K, Collins11}, the
formation of the first stars \citep{ABN02, 2007ApJ...654...66O,
2009Sci...325..601T, Xu08}, and the formation of stars in our own Galaxy
\citep{Collins11, 2011ApJ...727L..20K, Collins12a}.

### 2013-06-25 ms.tex
Many small changes to all files -- GB final edits

\begin{abstract}
This paper describes the open-source code \enzo, which uses 
block-structured adaptive mesh refinement to provide high spatial and
temporal resolution for modeling astrophysical fluid flows.  The code
is Cartesian, can be run in 1, 2, and 3 dimensions, and supports a
large variety of physics, including hydrodynamics, ideal and non-ideal
magnetohydrodynamics, N-body dynamics (and, more broadly, self-gravity
of fluids and particles), primordial gas chemistry, optically-thin
radiative cooling of primordial and metal-enriched plasmas (as well as
some optically-thick cooling models), radiation
transport, cosmological expansion, and models for star formation and
feedback in a cosmological context.  In addition to explaining the
algorithms implemented, we present solutions for a wide range of test
problems, demonstrate the code's parallel scaling properties, and
discuss the \enzo\ collaboration's code development methodology.
\end{abstract}

>>>>

\begin{abstract}
This paper describes the open-source code \enzo, which uses 
block-structured adaptive mesh refinement to provide high spatial and
temporal resolution for modeling astrophysical fluid flows.  The code
is Cartesian, can be run in 1, 2, and 3 dimensions, and supports a wide
variety of physics including hydrodynamics, ideal and non-ideal
magnetohydrodynamics, N-body dynamics (and, more broadly, self-gravity
of fluids and particles), primordial gas chemistry, optically-thin
radiative cooling of primordial and metal-enriched plasmas (as well as
some optically-thick cooling models), radiation
transport, cosmological expansion, and models for star formation and
feedback in a cosmological context.  In addition to explaining the
algorithms implemented, we present solutions for a wide range of test
problems, demonstrate the code's parallel performance, and
discuss the \enzo\ collaboration's code development methodology.
\end{abstract}

### 2013-06-25 numerical-amr.tex
Many small changes to all files -- GB final edits

In order to simplify bookeeping, a number of restrictions are imposed on the subgrids:
\begin{itemize}
 \item The refinement factor $r$, or the ratio of the coarse cell width to the fine cell width, must be an integer.  In practice, \enzo\ almost always uses $r=2$, since tests have shown this typically provides the best performance.
 \item Each subgrid must begin and end on the boundary of a coarse cell.  This means that the number of cells in a patch will be a multiple of the refinement factor.
 \item A subgrid must be completely enclosed by its parent.
\end{itemize}

@@

From the point of view of the hydro solver, each grid is solved as an independent computational fluid dynamics (CFD) problem, with Dirichlet boundary conditions stored in the ghost zones.  At the end of each timestep, cells on a given level fill ghost zones by interpolating from the parent and copying from neighbors on the same level.  Grids then correct the flux at the interface boundary between fine and coarse zones, and finally project their active zone data to the parent grid.

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime)
   dt = ComputeTimeStep(0)
   EvolveLevel(0, dt)
   Time = Time + dt
   CheckForOutput(Time)
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level upon which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level)
   SetBoundaryValues
   while (Time < ParentTime)
      dt = ComputeTimeStep(level)
      SolveHydroEquations(dt)
      SolveOtherEquations(dt)
      Time = Time + dt
      SetBoundaryValues
      EvolveLevel(level+1, dt)
      FluxCorrection
      Projection
      RebuildHierarchy(level+1)
\end{verbatim}

@@

In the first case, we allow for four types of boundary conditions
(represented here as being at $x=0$, with positive $x$ interior and
negative $x$ exterior to the domain):
\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a mirror, with the solution on the boundary reflecting the solution in the computational domain ($q(-x) = q(+x)$) but with the velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.}  We approximate outflow boundary conditions by duplicating the solution at the edge of the computational domain: $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a pre-determined function ($q(-x) = q0(-x,t)$).  The code provides a simple way to set inflow values that are constant in both time and space over an entire face, or hooks are provided for special cases where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

\item{\em Baryon mass.}  This refinement criterion is designed to
  mimic a Lagrangian method in that it tries to keep a fixed mass
  resolution.  In this method, a cell is refined if the baryonic mass
  in the cell ($M_g = \rho\, (\Delta x)^d$) is larger than a specified value:
\begin{equation}
M_g > \rho_{\rm flag}\, (\Delta x_{\rm root})^d r^{\epsilon_l l},
\end{equation}
where $\rho_{\rm flag}$ is the equivalent density on the root grid
required for refinement, $\Delta x_{\rm root}$ is the root grid cell
spacing, $r$ is the refinement factor, $l$ is the level and
$\epsilon_l$ is a parameter that can be used to make the refinement
more aggressive ($\epsilon_l < 0$) or less aggressive ($\epsilon_l >
0$) than pure Lagrangian refinement --- in other words,
super-Lagrangian or sub-Lagrangian in refinement.

@@

\item{\em Optical depth.} This refinement criterion, typically used for simulations using radiation transport, ensures that cells on the highest level of refinement have an \ion{H}{1} optical depth lower than unity.  The optical depth of a cell is estimated to be:
\begin{equation}
\tau = \sigma_{\rm HI} n_{\rm HI} \Delta x 
\end{equation}
where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width (which is assumed to be the same along each axis).  If $\tau > 1$, the cell is refined
further, up to the maximum level of refinement.

>>>>

In order to simplify bookkeeping, a number of restrictions are imposed on the subgrids:
\begin{itemize}
 \item The refinement factor $r$, or the ratio of the coarse cell width to the fine cell width, must be an integer.  In practice, we generally adopt $r=2$, since tests have shown this typically provides the best performance.
 \item Each subgrid must begin and end on the boundary of a coarse cell.  This means that the number of cells in a patch will be a multiple of the refinement factor.
 \item A subgrid must be completely enclosed by its parent.
\end{itemize}

@@

From the point of view of the hydro solver, each grid is solved as an independent computational fluid dynamics (CFD) problem, with Dirichlet boundary conditions stored in the ghost zones.  At the beginning of each timestep, cells on a given level fill ghost zones by interpolating from the parent and copying from neighbors on the same level.  Grids then correct the flux at the interface boundary between fine and coarse zones, and finally project their active zone data to the parent grid.

\begin{figure}
\begin{center}
%\includegraphics[width=0.6\textwidth]{figures/wcycle.eps}
\includegraphics[width=0.6\textwidth]{figures/timestepping.eps}
\end{center}
\caption{\emph{Left:} Example of the timesteps on a 2-level AMR
  hierarchy.  \enzo\ does not restrict the timesteps on the finer levels
  to be a factor of $1/2^n$ smaller.  \emph{Right:} The order in which
  the AMR grids are evaluated on each level.\vspace{1ex}}
\label{fig:wcycle}
\end{figure}

@@

The basic control algorithm looks very much like any used in a grid-based code of a single resolution, and can be written schematically as:
\begin{verbatim}
InitializeHierarchy
While (Time < StopTime):
   dt = ComputeTimeStep(0)
   EvolveLevel(0, dt)
   Time = Time + dt
   CheckForOutput(Time)
\end{verbatim}
The AMR control algorithm is contained within the recursive
EvolveLevel algorithm, which takes the level upon which to operate as an
argument.  This looks something like the following pseudo-code:
\begin{verbatim}
EvolveLevel(level):
   SetBoundaryValues
   while (Time < ParentTime):
      dt = ComputeTimeStep(level)
      SolveHydroEquations(dt)
      SolveOtherEquations(dt)
      Time = Time + dt
      SetBoundaryValues
      EvolveLevel(level+1, dt)
      FluxCorrection
      Projection
      RebuildHierarchy(level+1)
\end{verbatim}

@@

In the first case, we allow for four types of boundary conditions
(for simplicity, represented here as being at $x=0$, with positive $x$ interior and
negative $x$ exterior to the domain):
\begin{enumerate}
  \item{\em Reflecting.} In this case, the boundary behaves like a mirror, with the solution on the boundary reflecting the solution in the computational domain ($q(-x) = q(+x)$) but with the velocity normal to the boundary reversed: $v_x(-x) = -v_x(+x)$.
  \item{\em Outflow.}  We approximate outflow boundary conditions by duplicating the solution at the edge of the computational domain: $q(-x) = q(0)$.
  \item{\em Inflow.} In this case, the boundary values are fixed by a pre-determined function ($q(-x) = q0(-x,t)$).  The code provides a simple way to set inflow values that are constant in both time and space over an entire face, or hooks are provided for special cases where the boundary must be time-varying.
  \item{\em Periodic.} The boundary solution is obtained from the other side of the grid: $q(-x) = q(x_{\rm{max}}-x)$.
\end{enumerate}

@@

\item{\em Baryon mass.}  This refinement criterion is designed to
  mimic a Lagrangian method in that it tries to keep a fixed mass
  resolution.  In this method, a cell is refined if the baryonic mass
  in the cell ($M_g = \rho\, (\Delta x)^d$) is larger than a specified value:
\begin{equation}
M_g > \rho_{\rm flag}\, (\Delta x_{\rm root})^d r^{\epsilon_l l},
\end{equation}
where $\rho_{\rm flag}$ is the equivalent density on the root grid
required for refinement, $\Delta x_{\rm root}$ is the root grid cell
spacing, $r$ is the refinement factor, $l$ is the level and
$\epsilon_l$ is a parameter that can be used to make the refinement
more aggressive ($\epsilon_l < 0$) or less aggressive ($\epsilon_l >
0$) than pure Lagrangian refinement --- in other words,
super-Lagrangian or sub-Lagrangian refinement.

@@

\item{\em Optical depth.} This refinement criterion, typically used for simulations using radiation transport, 
flags cells for refinement if they have an \ion{H}{1} optical depth lower than unity.  The optical depth of a cell is estimated to be:
\begin{equation}
\tau = \sigma_{\rm HI} n_{\rm HI} \Delta x 
\end{equation}
where $\sigma_{\rm HI}$ is the neutral atomic hydrogen absorption
cross section, $n_{\rm HI}$ is the proper \ion{H}{1} number density,
and $\Delta x$ is the cell width (which is assumed to be the same along each axis).  If $\tau > 1$, the cell is refined
further, up to the maximum level of refinement.

### 2013-06-25 numerical-chemistry.tex
Many small changes to all files -- GB final edits

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by \citet{abel97} and \citet{anninos97}, but have
since been extended with updated reaction rates and the inclusion of
deuterium species \citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  \enzo\
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation (see Section~\ref{sec.num.cooling} for
more details).  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., decoupled and operator-split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that are faster than those of the gas dynamics.  The gamma used by \enzo\ to
compute the temperature of the gas from the energy and density characteristics
utilizes a variable gamma that includes effects of the rotational state of
molecular hydrogen, enabling it to vary from $5/3$ (fully-atomic) to $7/5$
(fully molecular).  This further coupling of the chemical and thermal states of
the gas underscores the need for coupled chemistry and radiative cooling
solutions.

@@

\enzo\ solves the rate equations using a method based on a semi-implicit
formulation in order to provide a stable, positive definite and first-order
accurate solution.  The update discretization splits chemical changes into
formation components and destruction components and updated with a mixed set of
time states, as described in \citet{anninos97}.  The formation components of
species $S_i$ are computed at the current subcycle time, where the contribution
of species $S_i$ to its own destruction components are computed at the updated
time; all other contributions to the destruction component are computed at the
current time.  This mixed state improves accuracy, ensures species values
are positive definite, and is equivalent to one Jacobi iteration of an implicit
Euler solve.  This technique is optimized by taking the chemical intermediaries
\Hm and \HHp, which have large rate coefficients and low concentrations, and
grouping them into a separate category of equations.  Due to their fast
reactions, these species are very sensitive to small changes in the more
abundant species and are at all times in astrophysical calculations close to
their equilibrium values.  Attempting to resolve their formation and destruction
times would necessitate extremely small timesteps, and while these species
provide necessary channels to facilitate the formation of molecular hydrogen at
low densities, they do not otherwise significantly influence the abundance of
species with higher concentrations.  Therefore, reactions governing these two
species can be decoupled from the rest of the network and treated independently
through analytic solutions for equilibrium values.  This allows a large speedup
in solution as the timestepping scheme is applied only to the slower 7- or
10-species network (depending on whether deuterium is included or not), which
will be much closer to the overall hydrodynamic timestep of the simulation.

@@

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling eventually breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$ in gas of primordial composition.  Beyond this point,
modifications to the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.   The formation
of molecular hydrogen as catalyzed by dust was recently added to \enzo\ to
enable studies of low-metallicity gas, as well as the inclusion of appropriate
timestepping criteria to account for the input of ionizing radiation.  Even
with these modifications, a completely correct description of the cooling of
primordial gas at very high densities requires some form of radiation
transport, which will greatly increase the cost of the simulations.
Furthermore, at very high densities, the stiffness of the molecular hydrogen
reaction rates may require better than a first-order accurate solution; as
such, the transition to this regime will likely necessitate a fully-implicit,
iterative solver.

>>>>

The primordial non-equilibrium chemistry routines used in \enzo\ were first
described by \citet{abel97} and \citet{anninos97}, but have
since been extended with updated reaction rates and the inclusion of
deuterium species \citep{McGreer2008, 2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12 total
species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp, and HD.  \enzo\
also computes the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as well as
several models for a metagalactic UV background that heat the gas via
photoionization and photodissociation (see Section~\ref{sec.num.cooling} for
more details).  The chemical and thermal states of the gas can be updated either
at the same hydrodynamical timestep (i.e., decoupled and operator-split) or
through the same subcycling system (i.e., a coupled chemical and thermal
system).  The default behavior of \enzo\ is to couple these two systems at
subcycles of the hydrodynamic timestep; this results in updates to both the
chemical and thermal states of the gas (which also inform the temperature, the
reaction rate coefficients and the cooling functions of the gas) on timescales
that are faster than those of the gas dynamics.  The gamma used by \enzo\ to
compute the temperature of the gas from the energy and density characteristics
utilizes a variable gamma that includes effects of the rotational state of
molecular hydrogen, enabling it to vary from $5/3$ (fully-atomic) to $7/5$
(fully molecular).  This further coupling of the chemical and thermal states of
the gas underscores the need for coupled chemistry and radiative cooling
solutions.

@@

\enzo\ solves the rate equations using a method based on a semi-implicit
formulation in order to provide a stable, positive definite and first-order
accurate solution.  The update discretization splits chemical changes into
formation components and destruction components and updates with a mixed set of
time states, as described in \citet{anninos97}.  The formation components of
species $S_i$ are computed at the current subcycle time, where the contribution
of species $S_i$ to its own destruction components are computed at the updated
time; all other contributions to the destruction component are computed at the
current time.  This mixed state improves accuracy, ensures species values
are positive definite, and is equivalent to one Jacobi iteration of an implicit
Euler solve.  This technique is optimized by taking the chemical intermediaries
\Hm and \HHp, which have large rate coefficients and low concentrations, and
grouping them into a separate category of equations.  Due to their fast
reactions, these species are very sensitive to small changes in the more
abundant species and are (at almost all times in astrophysical calculations) close to
their equilibrium values.  Attempting to resolve their formation and destruction
times would necessitate extremely small timesteps.
%and while these species
%provide necessary channels to facilitate the formation of molecular hydrogen at
%low densities, they do not otherwise significantly influence the abundance of
%species with higher concentrations.  
Therefore, reactions governing these two
species can be decoupled from the rest of the network and treated independently
through analytic solutions for equilibrium values.  This allows a significant speedup
in the solution speed, as the timestepping scheme is applied only to the slower 7- or
10-species network (depending on whether deuterium is included or not), which
will be much closer to the overall hydrodynamic timestep of the simulation.

@@

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references was only claimed to be valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling eventually breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$ in gas of primordial composition.  Beyond this point,
modifications to the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.   The formation
of molecular hydrogen as catalyzed by dust was recently added to \enzo\ to
enable studies of low-metallicity gas, as well as the inclusion of appropriate
timestepping criteria to account for the input of ionizing radiation.  Even
with these modifications, a completely correct description of the cooling of
primordial gas at very high densities requires some form of radiation
transport, which will greatly increase the cost of the simulations.
Furthermore, at very high densities, the stiffness of the molecular hydrogen
reaction rates may require better than a first-order accurate solution; as
such, the transition to this regime will likely necessitate a fully-implicit,
iterative solver.

### 2013-06-25 numerical-conduction.tex
Many small changes to all files -- GB final edits

\enzo\ implements the equations of isotropic heat conduction in a manner
similar to that of \citet{2007ApJ...664..135P}.  
The isotropic flux of heat is given by
\begin{equation}
\myvec{F} = -\kappa_{\rm sp} \grad T
\end{equation}
where $\kappa_{\rm sp}$ is the Spitzer conduction coefficient, $4.6 \times 10^{-7}$~T$^{5/2}$ erg
s$^{-1}$~cm$^{-1}$~K$^{-1}$ \citep{1962pfig.book.....S}.  In this
situation we are using a
value for the Coulomb logarithm, $\log \Lambda = 37.8$, that is
appropriate for the intracluster medium \citep{1988xrec.book.....S} --
in astrophysically-relevant, fully ionized plasmas this value varies
by no more than 50\% (see, e.g., Smith et al. 2013, submitted).
It is quite possible that the local heat flux computed in this way 
can become unphysically large in
the high-temperature, low-density cluster regime when using this
formulation; therefore, we take into account the saturation of the heat
flux \citep{1977ApJ...211..135C} at a maximum level of
\begin{equation}
F_{sat} \simeq 0.4 n_e k_b T \left( \frac{2 k_b T}{\pi m_e} \right)^{1/2}.
\end{equation}
To ensure a smooth transition between
the Spitzer and saturated regimes, we define an effective
conductivity using the formalism of \citet{1988xrec.book.....S}
\begin{equation}
\kappa_{eff} = \frac{\kappa_{\rm sp}}{1 + 4.2 \lambda_e / \ell_T} \; ,
\end{equation}
where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\grad T|$ is the characteristic length scale of the local
temperature gradient.  We also assume that the conductivity of the
plasma can be described in terms of an effective conductivity, which
can be expressed as a fraction f$_{\rm sp}$ of the Spitzer conductivity
(where f$_{\rm sp} \leq 1.0$ are considered physically realistic values).  This takes
into account physical processes below the resolution limit of the
simulation, such as tangled magnetic fields, that can suppress heat
transport.

Thermal conduction in a plasma can be strongly affected by the
presence of magnetic field lines, which may suppress heat
flow perpendicular to the magnetic field.  In that case, we allow for
heat transport only parallel to the magnetic field lines in
magnetohydrodynamic simulations.  Mathematically, this is treated as:
\begin{equation}
\myvec{F} = -\kappa_{\rm sp} \myvec{b} (\myvec{b} \cdot \grad T)
\end{equation}
where $\myvec{b}$ is the unit vector denoting the direction of the
magnetic field.  As with the isotropic thermal conduction, we allow a
multiplicative factor f$_{\rm sp}$ to take into account the possible
behavior of magnetic fields below the resolution limit of the
simulation.

>>>>

\enzo\ implements the equations of isotropic heat conduction in a manner
similar to that of \citet{2007ApJ...664..135P}.  
The isotropic flux of heat is given by equation~(\ref{eq:conduction}) and we
use a value for the Spitzer conduction coefficient, 
$\kappa_{\rm sp} = 4.6 \times 10^{-7}$~T$^{5/2}$ erg
s$^{-1}$~cm$^{-1}$~K$^{-1}$ \citep{1962pfig.book.....S}.  In this
situation we are using a
value for the Coulomb logarithm, $\log \Lambda = 37.8$, that is
appropriate for the intracluster medium \citep{1988xrec.book.....S} --
in astrophysically-relevant, fully ionized plasmas this value varies
by no more than 50\% (see, e.g., Smith et al. 2013, submitted).
It is quite possible that the local heat flux computed in this way 
can become unphysically large in
the high-temperature, low-density cluster regime when using this
formulation; therefore, we take into account the saturation of the heat
flux \citep{1977ApJ...211..135C} at a maximum level of
\begin{equation}
F_{sat} \simeq 0.4 n_e k_b T \left( \frac{2 k_b T}{\pi m_e} \right)^{1/2}.
\end{equation}
To ensure a smooth transition between
the Spitzer and saturated regimes, we define an effective
conductivity using the formalism of \citet{1988xrec.book.....S}
\begin{equation}
\kappa_{eff} = \frac{\kappa_{\rm cond}}{1 + 4.2 \lambda_e / \ell_T} \; ,
\end{equation}
where $\lambda_e$ is the electron mean free path and $\ell_T \equiv T
/ |\grad T|$ is the characteristic length scale of the local
temperature gradient.  We also assume that the conductivity of the
plasma can be described in terms of an effective conductivity, which
can be expressed as a fraction f$_{\rm sp}$ of the Spitzer conductivity
(where f$_{\rm sp} \leq 1.0$ are considered physically realistic values).  This takes
into account physical processes below the resolution limit of the
simulation, such as tangled magnetic fields, that can suppress heat
transport.

Thermal conduction in a plasma can be strongly affected by the
presence of magnetic field lines, which may suppress heat
flow perpendicular to the magnetic field.  In that case, we allow for
heat transport only parallel to the magnetic field lines in
magnetohydrodynamic simulations.  Mathematically, this is given
by equation~(\ref{eq:anisotropic_conduction}.
As with the isotropic thermal conduction, we allow a
multiplicative factor f$_{\rm sp}$ to take into account the possible
suppression of magnetic fields below the resolution limit of the
simulation.
